Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-23 joint_inf.tex
notation notation notation

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

@@

Let $\mathcal{Y}$ be the set of assignments of characters to tips of our tree, and $\Xi$ the set of assignments of characters to the internal nodes of the tree.
Both $\mathcal{Y}$ and $\Xi$ are finite sets.
Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
%EM Hm, this doesn't make sense compared to the above.
Without loss of generality, assume $\mathcal{Y}=\Xi=\{0,1\}$.
For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_n(\mathbf{y}; \tau, t) = \sum_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t).
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\mathbf{y}; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_n'(\mathbf{y};\tau, t) = \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\mathbf{y};\tau, t)
\end{equation}
while using $\hat{\boldsymbol\xi}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.
Since $\Xi$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\boldsymbol\xi}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\boldsymbol\xi}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

Since we have a finite character alphabet, there are a finite number of possibilities of assignments of characters to tips $y_i$ or internal nodes $\xi_i$.
These are commonly called ``site patterns'' in the literature.
%EM in general aren't a different number of internal site patterns than tip site patterns?
We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state patterns $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$; these are defined by the number of leaves and the character alphabet.
This enables us to group identical terms, reformulating the likelihood equation
\begin{align}
L_n'(\mathbf{y};\tau, t) &= \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) \\
    &= \max_{\boldsymbol\xi} \ \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t) \\
%EM Can you think a little about z_j? The indexing seems a little problematic now that we are thinking of everything as a function of y.
    &= \max_{\mathbf{z}} \ \prod_{j=1}^{q} \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})} \label{eq:site_pattern_likelihood}
\end{align}
where $n_j(\mathbf{y})$ is the number of observations in the sample where site pattern $s_j$ was seen in $\mathbf{y}$.
%EM It's a small point, but note that the argument below justified the equation above, which is a little upside down.
Note that since $\mathcal{S}$ and $\mathcal{Z}$ are also finite sets, the value of each factor $P(y_i, \xi_i | \tau, t)$ will be maximized at one or more values of $\xi_i$, though independent of $i$ given $j$.
This independence allows us to group factors by those where $\xi_i$ equals $z_j$ so that $P(s_j, z_j | \tau, t)$ is maximized.

@@

\subsection{The case of infinite data}

We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_n''(\mathbf{y};\mathbf{z},\tau,t) = \prod_{j=1}^q \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})}
$$
so that
$$
\max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \max_{\mathbf{z}} \ L_n''(\mathbf{y};\mathbf{z},\tau,t).
$$
Assume our $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot \log P(s_j, z_j | \tau, t) \\
  &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)].
\end{align}
In the limit of infinite data we have
$$
\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) \rightarrow \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)]
$$
and that, since $z_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
\begin{align}
\frac{1}{n} \log L_n'(\mathbf{y};\tau, t) &= \max_{\mathbf{z}} \ \frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \nonumber \\
    &\rightarrow \max_{z_1, \ldots, z_q} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
    &= \max_{\mathbf{z}} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
    &= \max_{\mathbf{z}} \ \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(s_j | \tau, t)\right\} \nonumber \\
    &\qquad + \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(z_j | s_j, \tau, t)\right\}. \label{eq:site_pattern_profile_likelihood_mean}
\end{align}

@@

For clarity, define \eqref{eq:site_pattern_profile_likelihood_mean} as
$$
J(\tau, t; \tau^*, t^*) = \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \max_{z_j} \ \log P(z_j | s_j, \tau, t)]
$$
where $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
J(\tau', \hat{t}_1; \tau^*, t^*) > J(\tau^*, \hat{t}_2; \tau^*, t^*)
\end{equation}
for $\tau'\neq\tau^*$ with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
Here, $\{p_x,p_y,p_z\}\in[0,1/2]^3$ and $\{x,y,z\}\in[0,1]^3$.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t^*)$&$P(s_j|\tau_2,t^*)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

>>>>

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

\input{commands}

@@

Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alignmentSpace^{\nSiteRows\times n}$ and corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\ancestralStateSpace^{\nAncestralStateRows\times n}$.
Without loss of generality, assume $\alignmentSpace=\ancestralStateSpace=\{0,1\}$.
For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_n(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_n(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_n'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\ancestralStateSpace$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
These are commonly called ``site patterns'' in the literature.
%EM in general aren't a different number of internal site patterns than tip site patterns?
%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing. Maybe writing out the example would help? I also changed from "pattern" to "category" which might make it a little more clear.
We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$; these are defined by the number of leaves and the character alphabet.
Corresponding to each of these site patterns will be an ancestral state category that is most likely given a particular likelihood.
Since $\alignmentSpace$ and $\ancestralStateSpace$ are finite sets, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
This independence allows us to group factors by those where $\ancestralStateColumn_i$ belongs to an ancestral state category $\ancestralStateCategory_j$ so that, if $\alignmentColumn_i=\sitePattern_j$ and $\ancestralStateColumn_i\in \ancestralStateCategory_j$, $P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)$ is maximized.

For ancestral state categories $\fullAncestralStateCategories=[\ancestralStateCategory_1,\ldots,\ancestralStateCategory_q]\in\ancestralStateCategorySpace$ we reformulate the likelihood equation
\begin{align}
L_n'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
%EM Can you think a little about z_j? The indexing seems a little problematic now that we are thinking of everything as a function of y.
                             &= \max_{\fullAncestralStateCategories} \ \prod_{j=1}^{q} \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $n_j(\fullAlignment)$ is the number of observations in the sample where site pattern $\sitePattern_j$ was seen in $\fullAlignment$.

@@

@@

We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^q \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)}
$$
so that
$$
\max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) =
    \max_{\fullAncestralStateCategories} \ L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t).
$$
Assume our $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot  \log P(\sitePattern_j, \ancestralStateCategory_j | \tau, t) \\
        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)].
\end{align}
In the limit of infinite data we have
$$
\frac{1}{n} \log L_n''(\fullAlignment;,\tau,t) \rightarrow \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)]
$$
and that, since $\ancestralStateCategory_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
\begin{align}
\frac{1}{n} \log L_n'(\fullAlignment;\fullAncestralStateCategories,\tau, t)
    &= \max_{\fullAncestralStateCategories} \ \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) \nonumber \\
    &\rightarrow \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
    &= \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
    &= \max_{\fullAncestralStateCategories} \ \left\{\sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot \left[\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)\right]\right\}. \label{eq:site_pattern_profile_likelihood_mean}
\end{align}
Define the the information quantity
$$
H_{\tau^*,t^*}(\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\sitePattern_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
\ell_{\tau^*,t^*}(\tau,t) = \max_{\fullAncestralStateCategories} \ \left\{H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t)\right\}.
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    H_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
H_{\tau^*,t^*}(\tau,t) \le H_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    H_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
Here, $\{p_x,p_y,p_w\}\in[0,1/2]^3$ and $\{x,y,w\}\in[0,1]^3$.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t^*)$&$P(\sitePattern_j|\tau_2,t^*)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t)$&$P(\sitePattern_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
    \hline
$\sitePattern_j$    &00                              &01\\
    \hline
0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
    \hline
    \multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
    \hline
$\sitePattern_j$    &00                              &01\\
    \hline
0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}

\subsubsection{Upper bound for Farris partial likelihood}
